Online Updating of Variational Bayes for Heterogeneous Forecasts of Vehicle Trajectory

### 2018-03-09 Cars/midcanPresentation.tex
updates

\documentclass[8pt]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\begin{frame}
\section{Framework}
\frametitle{Thesis Framework}
Focus: Two sets of data
\begin{itemize}
\item Stuff Here about the general themes
\item Use of exact inference to improve VB
\item Hierarchical models with exact inference
\item Some new unit with time constrained inference
\item Want to use MCMC info to improve VB
\end{itemize}
\end{frame}

\begin{frame}
\section{Inference}
\frametitle{Bayesian Inference}
Note: reframe discussion with two sets of data from previous slide
\begin{itemize}
\item The posterior distribution for parameters $\theta$, $p(\theta | \textbf{z})$, can be inferred with Markov Chain Monte Carlo (MCMC). MCMC:
\begin{enumerate}
\item Allows 'exact' inference as the number of iterations increases
\item Can be very slow
\end{enumerate}
\item We focus on settings where MCMC inference for the additional unit is too slow to be useful.
\item Instead, Variational Bayes (VB) is used. VB:
\begin{enumerate}
\item Approximates the posterior with an analytical alternative $q_{\lambda}(\theta | \textbf{z})$.
\item Chooses the distribution $q$ that minimises a divergence funciton
\item Is typically much faster than MCMC.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Kullback Leibler Divergence and Stochastic Optimisation}
\begin{itemize}
\item The typical divergence is the Kullback Leibler Divergence from $q$ to $p$.
\end{itemize}
\begin{equation}
\label{KL-def}
KL[q_{\lambda}(\theta | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta | \textbf{z})] = E_{q_{\lambda}(\theta | \textbf{z})} \left[ \log(q_{\lambda}(\theta | \textbf{z})) - \log(p(\theta | \textbf{z})) \right],
\end{equation}
\begin{itemize}
\item Monte-Carlo simulations are usually unavailable due to the $p(\theta | \textbf{z})$ term.
\end{itemize}
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta | \textbf{z})} \left[\log(p(\theta, \textbf{z})) - \log(q_{\lambda}(\theta | \textbf{z}))\right],
\end{equation}
\begin{itemize}
\item Stochastic Estimates are available through Monte-Carlo:
\end{itemize}
\begin{equation}
\label{ELBO-MC}
\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j}, \textbf{z})) - \log(q_{\lambda}(\theta_{j} | \textbf{z})) \right)
\end{equation}
where $\theta_{j} \sim q_{\lambda}(\theta | \textbf{z})$.
\end{frame}

\begin{frame}
\frametitle{Gradient Estimators}
\begin{itemize}
\item For most $p$ and $q$ only Stochastic Gradient Ascent is available for optimisation. Repeatedly apply:
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
\end{equation}
\item ELBO gradients can be estimated by the Score Estimator:
\begin{equation}
\label{scoreDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
\end{equation}
where $\theta_j \sim q_{\lambda}(\theta | \epsilon)$.
\item Or the Reparametersied Estimator, where $\theta = f(\epsilon, \lambda)$:
\begin{equation}
\label{rpDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta \theta}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda}, 
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{itemize}
\end{frame}

\begin{frame}
\section{Online Heterogenous Forecasting: Problem}
\subsection{Motivation}
\frametitle{Self-Driving Vehicles}
\begin{itemize}
\item Self-Driving Vehicles are a thing soon
\item The navigation system has three steps
\begin{enumerate}
\item Detect surrounding traffic
\item Forecast surrounding traffic
\item Navigate through safely
\end{enumerate}
\item Forecasting has not had much work
\end{itemize}
\end{frame}

@@

\begin{frame}
\subsection{Data}
\frametitle{Next Generation Simulation}
\begin{itemize}
\item This is a dataset created for macro traffic modelling
\item It records all of the cars all of the time
\item The road is curved
\item Before and After of road straightening project
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Forecasting in the framework}
\begin{itemize}
\item That is a terrible title.
\item Split the data into two sets: train/test
\item Cars observed beforehand, cars encountered while driving
\end{itemize}
\begin{figure}
\centering
\includegraphics[width = 0.95\textwidth]{problem}
\end{figure}
\end{frame}

@@

\begin{frame}
\subsection{Modelling}
\frametitle{Auto-regressive Models for Trajectory Forecasting}
\begin{itemize}
\item Evidence of dynamics behaviour
\item Modelling approachs / relationship to test train set
\item Motivate the use of the three different models
\end{itemize}
\end{frame}

@@

\begin{frame}
\begin{itemize}
\item That picture showing the IH / CH fits
\item Evidence of heterogeneity and the suitability of the mixture model
\end{itemize}
\end{frame}

\begin{frame}
\section{Updating Variational Bayes}
\frametitle{Updating Variational Bayes}
\begin{itemize}
\item Motivation
\begin{figure}[ht]
\centering
\includegraphics[width = 0.95\textwidth]{timeUpdate}
\end{figure}
\item Notation: $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ and $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Approximating Bayesian Updating}
\begin{equation}
\label{updatePost}
p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
\end{equation}
\begin{itemize}
\item This only works if the right hand side can be evaluated
\item VB Gradient Estimators requires the joint distribution
\begin{equation}
\label{updateJoint}
p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
\end{equation}
\item We will replace it with this
\begin{equation}
\label{ApproxJoint}
\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S}).
\end{equation}
\item Sources of error: How close $q$ is to $p$
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Updating Gradient Estimators}
The updating score estimator:
\begin{align}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
\end{align}
where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
The updating reparameterised estimator:
\begin{equation}
\label{rpUpdate}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{frame}

>>>>

\documentclass[11pt]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

@@

\begin{frame}
\section{Background}
\subsection{Motivation}
\frametitle{Motivation}
\begin{itemize}
\item Bayesian Inference in Forecasting:
\begin{enumerate}
\item Provides forecast densities for the variable of interest.
\item Encorporates parameter uncertainty.
\item Is easy to extend to hierarchical models.
\end{enumerate}
\vspace{3mm}
\item Bayesian Inference is only as good as its compuation:
\begin{enumerate}
\item Computation based on Markov Chain Monte Carlo (MCMC) is 'exact', but slow.
\item Computation based on Variational Bayes (VB) is fast, but approximate.
\end{enumerate}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item We focus on Hierarchical Models where:
\begin{enumerate}
\item Inference conditioned on existing units in the hierarchy is available.
\item Inference conditioned on additional units in a short time-frame is desired.
\end{enumerate}
\vspace{3mm}
\item We aim to infer, and forecast, additional units in an online setting.
\item We want to leverage the existing information to improve the forecasts.
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Hierarchical Models}
\frametitle{Hierarchical Models}
\begin{itemize}
\item Hierarchical Models are structured into three levels:
\begin{enumerate}
\item The likelihood: $p(z_i | \theta_i)$ for each unit $i$ in the model,
\item The prior: $p(\theta_i | \beta)$ where $\beta$ is shared across all $i$,
\item The hyper-prior: $p(\beta)$ is fixed.
\end{enumerate}
\vspace{5mm}
\item For $i = 1, 2, \dots, N$, we assume that the posterior distribution $$p(\beta, \theta_1, \dots, \theta_N | z_1, \dots, z_N)$$ is available.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Hierarchical Models}
\begin{itemize}
\item What about $i = N+1$? Using $\textbf{z}_{1:j} = \{z_i | i = 1, \dots, j\}$, $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1})\int_{\beta}p(\theta_{N+1} | \beta)p(\beta | \textbf{z}_{1:N})d\beta.$$
\item If there is a time constraint, we might use VB and approximate $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
\item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
\end{itemize}
\end{frame}

\begin{frame}
\subsection{Variational Bayes}
\frametitle{Variational Bayes and the Kullback-Leibler Divergence}
\begin{itemize}
\item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
\begin{align}
\label{KL-def}
KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \nonumber \\
&E_q \left[ \log(q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})) - \log(p(\theta_{N+1} | \textbf{z}_{1:N+1})) \right],
\end{align}
\item The expectation is intractable for most $p$ and $q$.
\item It typically cannot be estimated with Monte-Carlo simulations.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The Evidence Lower Bound}
\begin{itemize}
\item Instead look at the Evidence Lower Bound (ELBO), 
\begin{equation}
\label{ELBO}
\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],
\end{equation}
\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{j, N+1} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{j, N+1} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
\item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item For an exponential family $p$ and factorisable $q$ Mean Field Variational Bayes provides a fast method to the ELBO. 
\item We do not restrict $p$ and $q$ this way and instead use Stochastic Gradient Ascent.
\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
\begin{equation}
\label{gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
\end{equation}
\end{itemize}
\begin{figure}
\centering
\includegraphics[width = 0.5\textwidth]{gradientascent}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Stochastic Optimisation}
\begin{itemize}
\item ELBO gradients can be estimated by the Score Estimator:
\begin{equation}
\label{scoreDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
\end{equation}
where $\theta_j \sim q_{\lambda}(\theta | \epsilon)$.
\item Or the Reparametersied Estimator, where $\theta = f(\epsilon, \lambda)$:
\begin{equation}
\label{rpDeriv}
\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta \theta}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda}, 
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{itemize}
\end{frame}

\begin{frame}
\section{Online Heterogeneous Forecasting: Problem}
\subsection{Self-Driving Vehicles}
\frametitle{Self-Driving Vehicles}
\begin{itemize}
\item Self-Driving Vehicles are a thing soon
\item The navigation system has three steps
\begin{enumerate}
\item Detect surrounding traffic
\item Forecast surrounding traffic
\item Navigate through safely
\end{enumerate}
\item Forecasting has not had much work
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Forecasting Methods}
The literature forecasts vehicles in three ways:
\begin{enumerate}
\item Constant Behaviour.
\item Predict Events.
\item Provide Point Estimates.
\end{enumerate}
\end{frame}

@@

\begin{frame}
\subsection{The Next Generation Simulation Dataset}
\frametitle{Next Generation Simulation}
\begin{itemize}
\item This is a dataset created for macro traffic modelling
\item It records all of the cars all of the time
\item The road is curved
\item Before and After of road straightening project
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Forecasting in the framework}
\begin{itemize}
\item That is a terrible title.
\item Data is split into a training set and a test set:
\item Training vehicles are observed before driving,
\item Test vehicles are encountered while driving.
\end{itemize}
\begin{figure}
\centering
\includegraphics[width = \textwidth]{problem}
\end{figure}
\end{frame}

\begin{frame}
\subsection{Auto-regressive Models for Trajectory Forecasting}
\frametitle{Auto-regressive Models for Trajectory Forecasting}
\begin{figure}
\centering
\includegraphics[width = \textwidth]{pacf}
\end{figure}
\begin{itemize}
\item Dynamics are implemented in with bivariate AR models
\item There are three approaches:
\begin{enumerate}
\item The Homogeneous Model
\item The Independent Heterogeneous Model
\item The Clustered Heterogeneous Model
\end{enumerate}
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Heterogeneity in the Model Posteriors}
\begin{figure}
\centering
\includegraphics[width = \textwidth]{HierSingleKDE}
\end{figure}
\end{frame}

\begin{frame}
\section{Updating Variational Bayes}
\frametitle{Updating Variational Bayes}
\begin{itemize}
\item Inference for self-driving vehicles needs to be fast.
\item Inference cannot slow down as the amount of data increases.
\begin{figure}[ht]
\centering
\includegraphics[width = \textwidth]{timeUpdate}
\end{figure}
\item How do we go from $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ to $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$ for $S < T$?.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Approximating Bayesian Updating}
\begin{itemize}
\item The Bayesian Update Equation: 
\begin{equation}
\label{updatePost}
p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
\end{equation}
\item Computation requires evaluating the joint distribution,
\begin{equation}
\label{updateJoint}
p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
\end{equation}
\item We will approximate the Bayesian Update Equation in VB,
\begin{equation}
\label{ApproxJoint}
\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S}).
\end{equation}
\item Source of additional error: How close $q$ is to $p$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Updating Gradient Estimators}
\begin{itemize}
\item The updating score estimator:
\begin{align}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
\end{align}
where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
\item The updating reparameterised estimator:
\begin{equation}
\label{rpUpdate}
\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
\end{equation}
where $\epsilon_j \sim q(\epsilon)$.
\end{itemize}
\end{frame}
